---
title: "Assignment1"
author: "Sudip Sarkar"
date: "Monday, April 13, 2015"
output: html_document
---

##Question 1 - What is mean total number of steps taken per day?

**Load Library**
```{r}
library(ggplot2)

```

**Loading and preprocessing the data**

```{r}
mydata <- read.table(file = "activity.csv", header = TRUE, sep = ",")

```
**Question 1 - Calculate the Sum of steps per day and eleminating NA records**
```{r}

 Totalstepsperday <- setNames(
                          aggregate(
                                     steps~as.Date(date),
                                     mydata,
                                     sum,
                                     na.rm = TRUE
                                    ),
                           c("date","steps")
                         )
```
**Draw a plot using ggplot2**
```{r}
Totalstepshist <- ggplot(Totalstepsperday,aes(x=date,y=steps)) + 
  geom_bar(stat="identity") + 
  ggtitle("Total number of steps per day (source data)")

```

**Print the graph**

```{r}
print(Totalstepshist)
```

**Print the summary showing the Median and mean**
```{r}
summary(Totalstepsperday$steps)
```

##Question 2 - What is the average daily activity pattern?

**Calculate the Average Steps**
```{r}
avgPattern <- aggregate(steps~interval,mydata,mean,na.rm = TRUE)
```

**Plot the graph using ggplot2**
```{r}

plotavgPattern <- ggplot(avgPattern,aes(x=interval,y=steps)) + 
                  geom_line() 
```

**Print the graph**
```{r}
print(plotavgPattern)
```

**Finding the Step with the peak value for steps**
```{r}
with(avgPattern,avgPattern[steps == max(steps),])
```

##Question 3 - Inputing missing values

**Number of NA values in the data set**
```{r}
length(mydata[is.na(mydata)])
```

**Copy the Data in a new variable**
```{r}
newmydata <- mydata
```

**Use Loop to fill in average values wherever there is NA**
```{r}
    for (i in 1:nrow(newmydata)){
          if (is.na(newmydata$steps[i])){
            interval_val <- newmydata$interval[i]
            row_id <- which(avgPattern$interval == interval_val)
            steps_val <- avgPattern$steps[row_id]
            newmydata$steps[i] <- steps_val
          }
     }
```

**Calculate the Average** 
```{r}
Totalstepsperday_imputed <- setNames(
                          aggregate(
                                     steps~as.Date(date),
                                     newmydata,
                                     sum,   na.rm = TRUE
                                    ),
                           c("date","steps")
                         )
```

**Call ggplot to prepare and print graph**
```{r}
  Totalstepshist_imputed <- ggplot(Totalstepsperday_imputed,aes(x=date,y=steps)) + 
                            geom_bar(stat="identity") + 
  ggtitle("Total number of steps per day (Fixed data)")

  print(Totalstepshist_imputed)
```  

**Calculate mean**
```{r}
mean(Totalstepsperday_imputed$steps)
```

**Calculate Median**
```{r}
median(Totalstepsperday_imputed$steps)
```


**Comparing the imputed plot with the previous plot, it is observed that there is no major impactin estimating the total number of steps per day.**



##Question 4 - Are there differences in activity patterns between weekdays and weekends?

```{r}
mydata$date <- as.Date(mydata$date, "%Y-%m-%d")
```
```{r}
# add a new column to indicate day of the week 
mydata$day <- weekdays(mydata$date)

#add a new column called day type and initialize to weekday
mydata$day_type <- c("weekday")
 
# If day is Saturday or Sunday, Change day_type as weekend
for (i in 1:nrow(mydata)){
  if (mydata$day[i] == "Saturday" || mydata$day[i] == "Sunday"){
    mydata$day_type[i] <- "weekend"
  }
 }

# convert day_time from character to factor
mydata$day_type <- as.factor(mydata$day_type)

week_interval_steps <- aggregate(steps ~ interval+day_type, mydata, mean)

qplot(interval, steps, data=week_interval_steps, geom=c("line"), xlab="Interval", 
      ylab="Number of steps", main="") + facet_wrap(~ day_type, ncol=1)
```
